# Data visualisation with R

Let’s first introduce the data set that will be used throughout this chapter. The data set, `ames` is a part of the Ames Housing Price data, containing 165 observations and 12 features including the sale date and price.

```{r }
str(ames)
```

## Scatterplots

Are the bigger houses more expensive?  In base R, we can use the `plot()` function to generate this scatterplot with the first argument being the variable on the x-axis and the second argument being the variable on the y-axis.

```{r pressure, echo=FALSE}
plot(ames$Gr_Liv_Area, ames$Sale_Price)
```

It's OK but not nice.  The `plot()` function provides a rich capability of customization by setting the graphical parameters. 

```{r}
plot(ames$Gr_Liv_Area, ames$Sale_Price, 
     col = "red", 
     xlab = "Living Area", 
     ylab = "Sale Price", 
     main = "Sale Price vs. Living Area",  
     pch = 1, #Shape of the points
     cex = 1) #Size of text and symbols
```

Although the `plot()` function gets the work done, the `ggplot2` package provides a superior user experience. It's a member of the `tidyverse` package, you don’t need to install it separately if tidyverse was already installed. 

```{r}
library(ggplot2)
ggplot(data = ames) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, y = Sale_Price))
```

In a nutshell, `ggplot2` implements the grammar of graphics, a coherent system for describing and building graphs.  We always start with the function `ggplot()` with a data frame or tibble as its argument. To generate a scatterplot, you can add a layer using the `+` operator followed by the `geom_point()` function, which is one of the many available geoms in ggplot. Inside `geom_point()`, you need to set the value of the mapping argument. The mapping argument takes a functional form as `mapping = aes()`, where the `aes` is short for aesthetics. For example, you can use `aes()` to tell ggplot to use which variable on the x-axis, which variable on the y-axis. 

```{r}
ggplot(data = ames) + 
  geom_point(mapping = aes(Gr_Liv_Area, y = Sale_Price), 
             color = "red",
             shape = 2,
             size = 0.5,
             alpha = 0.3) #transparency level of the points
```

Suppose we want to use different colors according to the different `House_Style` in the scatterplot.

```{r}
ggplot(data = ames) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           color = House_Style), 
             size = 1)
```

In addition to color, you can also map a discrete variable to the size aesthetic.

```{r, warning=FALSE}
ggplot(data = ames) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           size = House_Style), 
             alpha = 0.5)
```

Or

```{r}
ggplot(data = ames) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           shape = Exter_Qual), 
             alpha = 0.5,
             size = 2)
```
  
and multiple mapping:

```{r}
ggplot(data = ames) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           shape = Exter_Qual,
                           color = Central_Air), 
             alpha = 0.5,
             size = 2)
```

## Smootlines

```{r}
ggplot(data = ames) + 
  geom_smooth(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price), 
              size = 1.5)
```

```{r}
ggplot(data = ames) + 
  geom_smooth(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price), 
              size = 1.5,
              method = "lm")
```

And 

```{r}
ggplot(data = ames) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price)) +
  geom_smooth(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price)) 

# Or with global mapping

ggplot(data = ames, 
       mapping = aes(x = Gr_Liv_Area, 
                     y = Sale_Price,
                     color = Central_Air)) +
  geom_point() +
  geom_smooth() 
```

## Interactive graphs

JavaScript is one of the most widely used language to create interactive webpages (html). There is an R [package](https://www.htmlwidgets.org), `htmlwidgets`, to bind R commands to various interactive JavaScript libraries that provides a great framework for graphs. The interactive components (“widgets”) created using the framework can be used at the R console, seamlessly embedded within R Markdown documents, Shiny web applications, saved as standalone web pages for ad-hoc sharing via email, Dropbox, etc.

There are a number of [widgets](http://gallery.htmlwidgets.org) already available, that you can install and easily make interactive visualizations. 

The `htmlwidgets`, by default, either run locally in your web browser or in the R Studio viewer. If you use R Markdown, the html pages rendered contain the full JavaScript code, so you can also also deploy them to a standard web server (like github pages).

Let's see one of those widgets, [`plotly`](https://plotly.com/r/), which binds R commands to o the JavaScript [`plotly.js`](https://plotly.com/javascript/) graphing library. The `plotly` package helps translate `ggplot2` graphics to an interactive web-based version.

```{r, warning=FALSE, message=FALSE}
# First install the package, if you haven't yet
library(plotly)

# our previous plot
p <- ggplot2::ggplot(data = ames, 
       mapping = aes(x = Gr_Liv_Area, 
                     y = Sale_Price,
                     color = Central_Air)) +
  geom_point() +
  geom_smooth()

# Converts ggplot2 to plotly
ggplotly(p)
```
## Shiny

`shiny` accepts user input.  Therefore, you can make the plot design by the user. Because it executes an actual R code, `shiny` requires its own server.  There are several ways to run an `shiny` app. A simple one is R command line.  Or you can call it from a R Markdown document. Or host it at `ShinyApps.io`.

Let's have a simple example:

```{r, message=FALSE, warning=FALSE}
library(shiny)
library(ggplot2)
library(dplyr)
library(RBootcamp)


ui <- fluidPage(    
  
  # Give the page a title
  titlePanel("Housing in Ames"),
  
  # Generate a row with a sidebar
  sidebarLayout(      
    
    # Define the sidebar with one input
    sidebarPanel(
      selectInput("neighborhood", "Neighborhood:", 
                  choices=unique(ames$Neighborhood)),
      hr(),
      helpText("Data from Ames Iowa Housing")
    ),
    
    # Create a spot for the barplot
    mainPanel(
      plotOutput("HousingStylePlot")  
    )
    
  )
)

## Server

# Define a server for the Shiny app
server <- function(input, output) {
  
  # Fill in the spot we created for a plot
  output$HousingStylePlot <- renderPlot({
    
    ames %>% 
      filter(Neighborhood == input$neighborhood) %>% 
      ggplot(aes(x = Gr_Liv_Area, 
                     y = Sale_Price,
                     color = Central_Air)) +
      geom_point() +
      geom_smooth()
  })
}

shinyApp(ui, server)
```



## Histograms & Density

How do we visualize continuous variables? One popular plot is called histograms.

```{r}
hist(amesdata$Sale_Price)
```

And density with ggplot

```{r}
ggplot(data = amesdata, aes(x = Sale_Price)) +
  geom_histogram(aes(y = ..density..)) +
  geom_density(color = "red",
               size = 1.5)
```

## Multiple plots

```{r}
fc <- ggplot(data = amesdata, aes(x = Sale_Price)) +
  geom_histogram(aes(y = ..density..)) +
  geom_density(color = "red",
               size = 1.5)
fc + facet_wrap("Street")
```
## Lables etc.

```{r}
fc + 
  xlab("Sale Price") + 
  ylab("Density") + 
  ggtitle("Price vs. Area") +
  theme(axis.text = element_text(size = 25, color = "red")) +
  theme(plot.title = element_text(size = 24, 
                                     color = "magenta", 
                                     face = "bold",
                                     hjust = 0.5)) 
```

## Add lines

Let's create some series (compounding series) of \$1000:

```{r}
t <- seq(0, 10, 1/4)
A1 <- 1000*(1+0.05)^t
A2 <- 1000*(1+0.03)^t
A3 <- 1000*(1+0.07)^t
A4 <- 1000*(1+0.1)^t
```

```{r}
plot(t, A1)
```

Here is a simple but beautiful plot ... So you may not want to use ggplot all the time:  

```{r}
plot(t, A1,  ylim = c(1000,2000), 
     type = "l", xlab = "Time (in years)", 
     ylab = "Accumulated Value (in $)", 
     main = "Accumulated Value of $1000 Investment", col = "black")
lines(t, A2, type = "l", col = "blue", lty = 2, lwd = 2)
lines(t, A3, type = "l", col = "red", lty = 3, lwd = 3)
legend("bottomright", 
       legend = c("3% Interest", "5% Interest", "7% Interest"), 
       col = c("blue", "black", "red"), 
       lty = c(2, 1, 3), bty = "o", cex = 0.75)
```

Or we can put them next to each other:

```{r}
par(mfrow=c(2,2), oma = c(0,0,2,0)) # puts 4 plots in one window (2x2)

plot(t, A1, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.05)")

plot(t, A2, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.03)")

plot(t, A3, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.07)")

plot(t, A4, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.1)")

mtext("Accumulated Value of $1000 Investment",
      outer=TRUE, cex = 1.5, col="olivedrab")
```

## Pairwise relationship

R package `corrplot` provides a visual exploratory tool on correlation matrix that supports automatic variable reordering to help detect hidden patterns among variables.  See more details [here](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)

```{r}
library(corrplot)
M = cor(mtcars)
corrplot(M, method = 'number') # colorful number
corrplot(M)
corrplot(M, order = 'AOE')
corrplot.mixed(M, order = 'AOE')
```

## Conditional Scatterplot

To visualize the relationship between two continuous variables but for different levels of a factor variable you can create a conditional scatterplot with `coplot()`. Note that we haven't converted `maturity.stage` to a factor variable.

```{r}
library(RBootcamp)
coplot(weight ~ DML | maturity.stage, data = squid1)
```

## `panel()`

To explore the relationships between multiple continuous variables we can have a pairs plot. 

```{r}
pairs(mtcars[, c(1, 3, 6)])
```

