# Data visualisation with R

Let’s first introduce the data set that will be used throughout this chapter. The data set is a part of the Ames Housing Price data, containing 165 observations and 12 features including the sale date and price.

```{r }
library(AmesHousing)
amesdata <- make_ames()
str(amesdata)
```

## Scatterplots

Are the bigger houses more expensive?  In base R, we can use the `plot()` function to generate this scatterplot with the first argument being the variable on the x-axis and the second argument being the variable on the y-axis.

```{r pressure, echo=FALSE}
plot(amesdata$Gr_Liv_Area, amesdata$Sale_Price)
```

It's OK but not nice.  The `plot()` function provides a rich capability of customization by setting the graphical parameters. 

```{r}
plot(amesdata$Gr_Liv_Area, amesdata$Sale_Price, 
     col = "red", 
     xlab = "Living Area", 
     ylab = "Sale Price", 
     main = "Sale Price vs. Living Area",  
     pch = 1, #Shape of the points
     cex = 1) #Size of text and symbols
```

Although the `plot()` function gets the work done, the `ggplot2` package provides a superior user experience. It's a member of the `tidyverse` package, you don’t need to install it separately if tidyverse was already installed. 

```{r}
library(ggplot2)
ggplot(data = amesdata) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, y = Sale_Price))
```

In a nutshell, `ggplot2` implements the grammar of graphics, a coherent system for describing and building graphs.  We always start with the function `ggplot()` with a data frame or tibble as its argument. To generate a scatterplot, you can add a layer using the `+` operator followed by the `geom_point()` function, which is one of the many available geoms in ggplot. Inside `geom_point()`, you need to set the value of the mapping argument. The mapping argument takes a functional form as `mapping = aes()`, where the `aes` is short for aesthetics. For example, you can use `aes()` to tell ggplot to use which variable on the x-axis, which variable on the y-axis. 

```{r}
ggplot(data = amesdata) + 
  geom_point(mapping = aes(Gr_Liv_Area, y = Sale_Price), 
             color = "red",
             shape = 2,
             size = 0.5,
             alpha = 0.3) #transparency level of the points
```

Suppose we want to use different colors according to the different `House_Style` in the scatterplot.

```{r}
ggplot(data = amesdata) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           color = House_Style), 
             size = 1)
```

In addition to color, you can also map a discrete variable to the size aesthetic.

```{r, warning=FALSE}
ggplot(data = amesdata) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           size = House_Style), 
             alpha = 0.5)
```

Or

```{r}
ggplot(data = amesdata) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           shape = Exter_Qual), 
             alpha = 0.5,
             size = 2)
```
  
and multiple mapping:

```{r}
ggplot(data = amesdata) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price, 
                           shape = Exter_Qual,
                           color = Central_Air), 
             alpha = 0.5,
             size = 2)
```

## Smootlines

```{r}
ggplot(data = amesdata) + 
  geom_smooth(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price), 
              size = 1.5)
```

```{r}
ggplot(data = amesdata) + 
  geom_smooth(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price), 
              size = 1.5,
              method = "lm")
```

And 

```{r}
ggplot(data = amesdata) + 
  geom_point(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price)) +
  geom_smooth(mapping = aes(x = Gr_Liv_Area, 
                           y = Sale_Price)) 

# Or with global mapping

ggplot(data = amesdata, 
       mapping = aes(x = Gr_Liv_Area, 
                     y = Sale_Price,
                     color = Central_Air)) +
  geom_point() +
  geom_smooth() 
```

## Histograms & Density

How do we visualize continuous variables? One popular plot is called histograms.

```{r}
hist(amesdata$Sale_Price)
```

And density with ggplot

```{r}
ggplot(data = amesdata, aes(x = Sale_Price)) +
  geom_histogram(aes(y = ..density..)) +
  geom_density(color = "red",
               size = 1.5)
```

## Multiple plots

```{r}
fc <- ggplot(data = amesdata, aes(x = Sale_Price)) +
  geom_histogram(aes(y = ..density..)) +
  geom_density(color = "red",
               size = 1.5)
fc + facet_wrap("Street")
```
## Lables etc.

```{r}
fc + 
  xlab("Sale Price") + 
  ylab("Density") + 
  ggtitle("Price vs. Area") +
  theme(axis.text = element_text(size = 25, color = "red")) +
  theme(plot.title = element_text(size = 24, 
                                     color = "magenta", 
                                     face = "bold",
                                     hjust = 0.5)) 
```

## Add lines

Let's create some series (compounding series) of \$1000:

```{r}
t <- seq(0, 10, 1/4)
A1 <- 1000*(1+0.05)^t
A2 <- 1000*(1+0.03)^t
A3 <- 1000*(1+0.07)^t
A4 <- 1000*(1+0.1)^t
```

```{r}
plot(t, A1)
```

Here is a simple but beautiful plot ... So you may not want to use ggplot all the time:  

```{r}
plot(t, A1,  ylim = c(1000,2000), 
     type = "l", xlab = "Time (in years)", 
     ylab = "Accumulated Value (in $)", 
     main = "Accumulated Value of $1000 Investment", col = "black")
lines(t, A2, type = "l", col = "blue", lty = 2, lwd = 2)
lines(t, A3, type = "l", col = "red", lty = 3, lwd = 3)
legend("bottomright", 
       legend = c("3% Interest", "5% Interest", "7% Interest"), 
       col = c("blue", "black", "red"), 
       lty = c(2, 1, 3), bty = "o", cex = 0.75)
```

Or we can put them next to each other:

```{r}
par(mfrow=c(2,2), oma = c(0,0,2,0)) # puts 4 plots in one window (2x2)

plot(t, A1, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.05)")

plot(t, A2, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.03)")

plot(t, A3, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.07)")

plot(t, A4, type = "l", xlab = "Time (in years)", 
     ylab = "AV - $", ylim = c(1000, 2500),
     main = "(i = 0.1)")

mtext("Accumulated Value of $1000 Investment",
      outer=TRUE, cex = 1.5, col="olivedrab")
```

## Pairwise relationship

R package `corrplot` provides a visual exploratory tool on correlation matrix that supports automatic variable reordering to help detect hidden patterns among variables.  See more details [here](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)

```{r}
library(corrplot)
M = cor(mtcars)
corrplot(M, method = 'number') # colorful number
corrplot(M)
corrplot(M, order = 'AOE')
corrplot.mixed(M, order = 'AOE')
```

