# Data Management

We will use the data set, **Ames Housing Price** data, from the `AmesHousing` package, containing 2930 observations and 81 features including the sale date and price.  It's a `tibble` but a dataframe version (`Rbootcamp::ames`) is also available in our package, `Rbootcamp`. And, we will introduce the `dplyr` package in many applications. See more about `dplyr` [here](https://dplyr.tidyverse.org/articles/dplyr.html).

```{r, warning=FALSE, message=FALSE}
library(AmesHousing)
library(dplyr)
amesdata <- make_ames()
glimpse(amesdata)
```

## Filter

Suppose we want to find the houses that are sold in Jan 2010. You can use the function `filter()` in the `dplyr` package, a member of the tidyverse package. We can use subsetting operations.  

```{r}
amesdata[amesdata$Year_Sold == 2010 & amesdata$Mo_Sold == 1, ] 
```

Or we can use `filter()`:

```{r}
library(dplyr)
dplyr::filter(amesdata, Year_Sold == 2010, Mo_Sold == 1)
```

## Arrange

Let's find the 10 houses with the highest sale prices by year.

```{r}
ar <- arrange(amesdata, Year_Sold, desc(Sale_Price))
ar
```

Or

```{r}
arr <- amesdata[order(amesdata$Year_Sold, desc(amesdata$Sale_Price)), ]
head(arr)
```

## Pipe

Pipes help us apply multiple operations sequentially on a given data.

```{r}
library(ggplot2)
amesdata %>% 
  filter(Year_Sold == 2009, Mo_Sold == 1) %>% 
  arrange(Year_Built) %>%
  ggplot(mapping = aes(x = Lot_Area, y = Sale_Price)) + 
  geom_point()
```

## Select

How do we select variables based on some characteristics

```{r}
amesdata %>% select(starts_with("Year"), Sale_Price)
amesdata %>% select(contains("Mo"))

```

## Create & `group_by()`

We may want to create new variables as functions of the existing ones by `mutate()`:

```{r, warning=FALSE, message=FALSE}
library(r02pro)
library(tidyverse)
amesdata %>% 
  select(Overall_Qual, Lot_Area, Sale_Price) %>%
  mutate(ave_price = Sale_Price/Lot_Area) 
```

Can we summarize by groups? First let's see an example for `summarize`:

```{r}
amesdata %>% 
  summarize(n_houses = n(),
            ave_liv_area = mean(Lot_Area),
            prob = c(0.25, 0.75),
            q_price = quantile(Sale_Price, c(0.25, 0.75),
                               na.rm = TRUE))
```

Another

```{r}
amesdata %>% 
  group_by(Overall_Qual) %>% 
  summarize(n_houses = n(),
            ave_liv_area = mean(Lot_Area),
            ave_price = mean(Sale_Price),
            na.rm = TRUE)
  
```

## More tools

### `subset()`

```{r, warning=FALSE, message=FALSE}
any(is.na(amesdata))

#Pay attention to subset().  This will be a time-saver
sub <- subset(amesdata, amesdata$Overall_Qual != "Fair")
dim(sub)
dim(amesdata)

#You can drop columns (variables) as well
amesless = subset(amesdata, select = c("Sale_Type", "Mo_Sold"))
head(amesless)
```

However, look at the `help(subset)`: "This is a convenience function intended for use interactively. *For programming it is better to use the standard subsetting functions like `[ ]`*, and in particular the non-standard evaluation of argument subset can have unanticipated consequences".

### `within()` & `with()`

Here is an example to use `within()`:

```{r}
ana <- within(amesdata, Sale_Price[Fence != 2] <- 0) 

#which is a short cut of

amesdata$Sale_Price[amesdata$Fence != 2] <- 0
```

And `with()`

```{r}
mean(with(amesdata, Sale_Price[Mo_Sold == 5  &  Overall_Qual =="Good"]))
```

### `aggregate()`

The `aggregate()` function in R can be used to calculate summary statistics for a dataset.

```{r}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'B', 'B', 'B'),
                 position=c('G', 'G', 'F', 'G', 'F', 'F'),
                 points=c(99, 90, 86, 88, 95, 99),
                 assists=c(33, 28, 31, 39, 34, 23),
                 rebounds=c(30, 28, 24, 24, 28, 33))

df
```

```{r}
#find mean points by team
aggregate(df$points, by=list(df$team), FUN=mean)
aggregate(df$points, by=list(df$team, df$position), FUN=mean)
```

We can also define our own functions

```{r}
mine <- function(x){
  return(sort(x))
}

aggregate(df$points, by=list(df$assists), FUN=mine)
```

## Tables

Here, we'll look at two-way tables.

### From Data with `table()`

Most data tables use categorical variables. Here is definition of input from `table()`:

> One or more objects which can be interpreted as factors (including numbers or character strings), or a list (such as a data frame) whose components can be so interpreted.

```{r}
tb1 <- table(amesdata$Lot_Shape, amesdata$House_Style)
tb1
```

```{r}
tb2 <- table(amesdata$Lot_Shape, amesdata$House_Style, amesdata$Street)
tb2
```

See what happens if you a continuous variable `amesdata$Lot_Area`:

```{r}
#tb2 <- table(amesdata$Lot_Shape, amesdata$Lot_Area, amesdata$Street)
#tb2
```

### `datatable()`

If we have more columns:

```{r}
DT::datatable(amesdata, rownames = FALSE, filter="top", options = list(pageLength = 10, scrollX=T) )
```

### With `describr`

The package [`describr`](https://cran.r-project.org/web/packages/descriptr/descriptr.pdf) has several good functions

```{r}
library(descriptr)
ds_screener(mtcars)
```

One continuous one factor:

```{r}
mtcars$cyl <- as.factor(mtcars$cyl)
ds_group_summary(mtcars, cyl, mpg)
```

### `xtabs()` & `ftabs()`

The `xtabs()`  & `ftabs()` functions allows you to quickly calculate frequencies for more factor variables.

```{r}
library(RBootcamp)
nt <- ftable(xtabs(~ year + maturity.stage + month, data = squid1))
nt
```

### Table with manual entry

```{r}
smoke <- matrix(c(51,43,22,92,28,21,68,22,9),ncol=3,byrow=TRUE)
colnames(smoke) <- c("High","Low","Middle")
rownames(smoke) <- c("current","former","never")
smoke <- as.table(smoke)
smoke
```

### Tools for tables

```{r}
margin.table(smoke, 1)
margin.table(smoke, 2)
smoke/margin.table(smoke)
prop.table(smoke)
```

```{r}
mosaicplot(smoke,main="Smokers",xlab="Status",ylab="Economic Class")
```

## `merge()`

The `merge()` function merges two data frames by common columns or row names, or do other versions of database join operations.  Here is an example:
  
```{r}
df1 = data.frame(StudentId = c(1:6), 
                 Marks = c("70", "84", "90", "93", "80", "76"))

df2 = data.frame(StudentId = c(2, 4, 6, 7, 8), 
                 city = c("Lahore", "Karachi", "Peshawar", "Quetta", "Multan")) 

df3 = merge(df1, df2, by = "StudentId")
df3

df4 = merge(df1, df2, by = "StudentId", all = TRUE)
df4
```

See the options of `merge()` by `?merge`.   We can do the same operation in `dplyr`:  

```{r}
library(dplyr)
df3= df1 %>% inner_join(df2,by="StudentId")
df3
```

