# Data Management

We will use the same data set, Ames Housing Price data, containing 165 observations and 12 features including the sale date and price. Suppose we want to find the houses that are sold in Jan 2010. You can use the function `filter()` in the `dplyr` package, a member of the tidyverse package. 

See more about `dplyr` [here](https://dplyr.tidyverse.org/articles/dplyr.html)

```{r }
library(AmesHousing)
amesdata <- make_ames()
```

## Filter

We can use subsetting operations

```{r}
amesdata[amesdata$Year_Sold == 2010 & amesdata$Mo_Sold == 1, ] 
```
  
Or we can use `filter()`:  

```{r}
library(dplyr)
dplyr::filter(amesdata, Year_Sold == 2010, Mo_Sold == 1)
```

## Arrange

Let's find the 10 houses with the highest sale prices by year

```{r}
ar <- arrange(amesdata, Year_Sold, desc(Sale_Price))
ar
```

Or

```{r}
arr <- amesdata[order(amesdata$Year_Sold, desc(amesdata$Sale_Price)), ]
head(arr)
```

## Pipe

Pipes help us apply multiple operations sequentially on a given data.

```{r}
library(ggplot2)
amesdata %>% 
  filter(Year_Sold == 2009, Mo_Sold == 1) %>% 
  arrange(Year_Built) %>%
  ggplot(mapping = aes(x = Lot_Area, y = Sale_Price)) + 
  geom_point()
```

## Select

How do we select variables based on some characteristics

```{r}
amesdata %>% select(starts_with("Year"), Sale_Price)
amesdata %>% select(contains("Mo"))

```

## Create & `group_by()`

We may want to create new variables as functions of the existing ones by `mutate()`:

```{r, warning=FALSE, message=FALSE}
library(r02pro)
library(tidyverse)
amesdata %>% 
  select(Overall_Qual, Lot_Area, Sale_Price) %>%
  mutate(ave_price = Sale_Price/Lot_Area) 
```
Can we summarize by groups?  First let's see an example for `summarize`:

```{r}
amesdata %>% 
  summarize(n_houses = n(),
            ave_liv_area = mean(Lot_Area),
            prob = c(0.25, 0.75),
            q_price = quantile(Sale_Price, c(0.25, 0.75),
                               na.rm = TRUE))
```

Another

```{r}
amesdata %>% 
  group_by(Overall_Qual) %>% 
  summarize(n_houses = n(),
            ave_liv_area = mean(Lot_Area),
            ave_price = mean(Sale_Price),
            na.rm = TRUE)
  
```

## More tools

### `subset()`

```{r, warning=FALSE, message=FALSE}
any(is.na(amesdata))

#Pay attention to subset().  This will be a time-saver
sub <- subset(amesdata, amesdata$Overall_Qual != "Fair")
dim(sub)
dim(amesdata)

#You can drop columns (variables) as well
amesless = subset(amesdata, select = c("Sale_Type", "Mo_Sold"))
head(amesless)
```

However, look at the `help(subset)`: "This is a convenience function intended for use interactively. *For programming it is better to use the standard subsetting functions like []*, and in particular the non-standard evaluation of argument subset can have unanticipated consequences".

### `within()` & `with()`

Here is an example to use `within()`:
  
```{r}
ana <- within(amesdata, Sale_Price[Fence != 2] <- 0) 

#which is a short cut of

amesdata$Sale_Price[amesdata$Fence != 2] <- 0
```


And `with()`

```{r}
mean(with(amesdata, Sale_Price[Mo_Sold == 5  &  Overall_Qual =="Good"]))
```

### `aggregate()`

The `aggregate()` function in R can be used to calculate summary statistics for a dataset.

```{r}
#create data frame
df <- data.frame(team=c('A', 'A', 'A', 'B', 'B', 'B'),
                 position=c('G', 'G', 'F', 'G', 'F', 'F'),
                 points=c(99, 90, 86, 88, 95, 99),
                 assists=c(33, 28, 31, 39, 34, 23),
                 rebounds=c(30, 28, 24, 24, 28, 33))

df
```

```{r}
#find mean points by team
aggregate(df$points, by=list(df$team), FUN=mean)
aggregate(df$points, by=list(df$team, df$position), FUN=mean)
```

We can also define our own functions

```{r}
mine <- function(x){
  return(sort(x))
}

aggregate(df$points, by=list(df$assists), FUN=mine)
```

